---
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
library(tidyverse)
library(gapminder)
# text width is 4.6 inch, but markdown scales down nicely even if we include widths and heights greater than 4.6
knitr::opts_chunk$set(fig.width = 2.8, fig.height = 2.8, cache = TRUE)
```

# Different types of plots

> What I can not create, I do not understand.  
> Richard Feynman


There are a few different plotting packages in R, but the most elegant and versatile one is `ggplot2`^[The name of the package is `ggplot2`, but the function is called `ggplot()`. For everything you've ever wanted to know about the grammar of graphics in R, see @wickham2016 .].
**gg** stands for **g**rammar of **g**raphics which means that we can make a plot by describing it one component at a time. 
In other words, we build a plot by adding layers to it.

This does not have to be many layers, the simplest `ggplot()` consist of just two components:

* the variables to be plotted, 
* a geometrical object (e.g., point, line, bar, box, etc.).

`ggplot()` calls geometrical objects **geoms**.


Figure \@ref(fig:chap04-fig-steps) shows some example steps for building a scatter plot, including changing its appearance ('theme') and faceting - an efficient way of creating separate plots for subgroups.


```{r, message=F, include = FALSE}
library(tidyverse)
library(gapminder)

mydata = gapminder

glimpse(mydata)
mydata$year %>% unique()

gapminder2007 = gapminder %>% 
  filter(year == 2007) %>% 
  mutate(gdpPercap_thousands = gdpPercap/1000)

```


```{r chap04-fig-steps, fig.height=10, fig.width=6, echo = FALSE, fig.cap = "Example steps for building and modiyfing a ggplot. (1) initialising the canvas, defining variables, (2) adding points, (3) colouring points by continent, (4) changing point type, (5) faceting, (6) changing the plot theme and the scale of the x variable."}
library(patchwork)

step1 = gapminder2007 %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp))

step2 = gapminder2007 %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point()

step3 = gapminder2007 %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp, colour = continent)) +
  geom_point()   +
  theme(legend.position = "none")

step4 = gapminder2007 %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp, colour = continent)) +
  geom_point(shape = 1)

step5 = gapminder2007 %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp, colour = continent)) +
  geom_point(shape = 1) +
  facet_wrap(~continent)

step5 = gapminder2007 %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp, colour = continent)) +
  geom_point(shape = 1) +
  facet_wrap(~continent)

step6 = gapminder2007 %>% 
  ggplot(aes(x = gdpPercap/1000, y = lifeExp, colour = continent)) +
  geom_point(shape = 1) +
  facet_wrap(~continent) +
  theme_light()

(step1 | step2) / (step3 | step4) / step5 / step6 + plot_annotation(tag_levels = "1", tag_prefix = "(", tag_suffix = ")")

```

\clearpage 

## Data - gapminder {#chap04-data}

We are using the gapminder dataset (https://www.gapminder.org/data) that has been put into an R package by @bryan2017 so we can load it with `library(gapminder)`.

```{r, message=F, include = TRUE}
library(tidyverse)
library(gapminder)

mydata = gapminder

glimpse(gapminder)
```



The dataset includes `r nrow(gapminder)` observations (rows) of `r ncol(gapminder)` variables (columns: `r colnames(gapminder)`).
`country`, `continent`, and `year` could be though of as grouping variables, whereas lifeExp (life expectancy), pop (population), and gdpPercap (Gross Domestic Product per capita) are values.

The years in this dataset span `r range(gapminder$year) %>% paste(collapse = " to ")` with 5-year intervals (so a total of `r gapminder$year %>% n_distinct()` different years).
It includes `r gapminder$country %>% n_distinct()` countries from `r gapminder$continent %>% n_distinct()` continents (`r gapminder$continent %>% unique()`).

The way for you to double check that all of the numbers we've quoted above are correct, run these lines:

```{r, results = "hide"}
library(tidyverse)
library(gapminder)
gapminder$year %>% unique()
gapminder$country %>% n_distinct()
gapminder$continent %>% unique()
```

Let's create a new shorter tibble called `gapminder2007` that only includes data for the year 2007.

```{r}
gapminder2007 = gapminder %>% 
  filter(year == 2007)

gapminder2007
```


## Anatomy of ggplot explained {#chap04-gganatomy}

We will now explain the six steps shown in Figure \@ref(fig:chap04-fig-steps).
Note that you only need the first two to make a plot, the rest are just to show you further functionality and optional customisations.


**(1)** Start by defining the variables, e.g., `ggplot(aes(x = var1, y = var2))`:

```{r, fig.keep = 'none'}
gapminder2007 %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp))
```

This creates the first plot in Figure \@ref(fig:chap04-fig-steps).

Although the above code is equivalent to:

```{r, fig.keep = 'none'}
ggplot(gapminder2007, aes(x = gdpPercap, y = lifeExp))
```

we tend to put the data first and then use the pipe (`%>%`) to send it to the `ggplot()` function.
This becomes useful when we add further data wrangling functions between the data and the `ggplot()`.
For example, our plotting pipelines often look like this:

```{r, fig.keep = 'none', eval = FALSE}
data %>% 
  filter(...) %>% 
  mutate(...) %>% 
  ggplot(aes(...)) +
  ...
```

The lines that come before the `ggplot()` function are piped, whereas from `ggplot()` onwards you have to use +.
This is because we are now adding different layers and customisations to the same plot.

`aes()` stands for **aes**thetics - things we can see. Variables are always inside the `aes()` function, which in return is inside a `ggplot()`.
Take a moment to appreciate the double closing brackets `))` - the first one belongs to `aes()`, the second one to `ggplot()`. 


**(2)** Choose and add a geometrical object

Let's ask `ggplot()` to draw a point for each observation by adding `geom_point()`:

```{r, fig.keep = 'none'}
gapminder2007 %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```

We have now created the second plot in Figure \@ref(fig:chap04-fig-steps), a scatter plot.

If we copy the above code and change just one thing - the `x` variable from `gdpPercap` to `continent` (which is a categorical variable) - we get what's called a strip plot.
This means we are now plotting a continuous variable (`lifeExp`) against a categorical one (`continent`).
But the thing to note is that the rest of the code stays exactly the same, all we did was change the `x = `.


```{r chap04-fig-stripplot, fig.width=3, fig.cap = "A stip plot using `geom_point()`."}
gapminder2007 %>% 
  ggplot(aes(x = continent, y = lifeExp)) +
  geom_point()
```

**(3)** specifying further variables inside `aes()`

Going back to the scatter plot (`lifeExp` vs `gdpPercap`), let's use `continent` to give the points some colour.
We can do this by adding `colour = continent` inside the `aes()`:

```{r, fig.width=4, fig.keep = 'none'}
gapminder2007 %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp, colour = continent)) +
  geom_point()
```

This creates the third plot in Figure \@ref(fig:chap04-fig-steps). It uses the default colour scheme and will automatically include a legend.
Still with just two lines of code (`ggplot(...)` + `geom_point()`).

**(4)** specifying aesthetics outside `aes()`

It is very important to understand the difference between including `ggplot` arguments inside or outside of the `aes()` function.

The main aesthetics (things we can see) are: **x**, **y**, **colour**, **fill**, **shape**, **size**, and any of these could appear inside or outside the `aes()` function.
Press F1 on, e.g., `geom_point()`, to see see the full list of aesthetics that can be used with this geom (this opens the Help tab).

Variables  (so columns of your dataset) have to be defined inside `aes()`. Whereas to apply a modification on everything, we can set an aesthetic to constant value outside of `aes()`.

For example, Figure \@ref(fig:chap04-fig-shapes) shows a selection of the point shapes built into R. The default shape used by `geom_point()` is number 16.

```{r chap04-fig-shapes, echo = FALSE, fig.width = 6,  fig.cap="A selection of shapes for plotting. Shapes 0, 1, and 2 are hollow, whereas for shapes 21, 22, and 23 we can define both a colour and a fill (for thes shapes, colour is the border around the fill).", warning = FALSE}

shapes = tibble(shape = c(0, 1, 2, 4, 8, 15, 16, 17, 21, 22, 23), x = 0:10)

ggplot(shapes, aes(x, y = 1)) + 
  geom_text(aes(label = shape), hjust = 0, nudge_x = -0.55) +
  geom_point(aes(shape = shape), size = 5, fill = "#2171b5") +
  scale_shape_identity() +
  theme_void()
```


To make all of the points in our figure hollow, let's set their shape to 1.
We do this by adding `shape = 1` inside the `geom_point()`:

```{r, fig.keep = 'none'}

gapminder2007 %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp, colour = continent)) +
  geom_point(shape = 1)

```

This creates the fourth plot in Figure \@ref(fig:chap04-fig-steps).

**(5)** From one plot to multiple with a single extra line

Faceting is a way to efficiently create the same plot for subgroups within the dataset.
For example, we can separate each continent on its own facet by adding `facet_wrap(~continent)` to our plot:


```{r fig.height=3.5, fig.width=7, fig.keep = 'none'}

gapminder2007 %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp, colour = continent)) +
  geom_point(shape = 1) +
  facet_wrap(~continent)

```

This creates the fifth plot in Figure \@ref(fig:chap04-fig-steps).
Note that we have to use the tilde (~) in `facet_wrap()`.
There is a similar function called `facet_grid()` that will create a grid of plots based on two grouping variables, e.g, `facet_grid(var1~var2)`.
Furthermore, facets are happy to quickly separate data based on a condition (so something you would usually use in a filter).

```{r chap04-fig-facetcond, fig.width=6, fig.cap = "Using a filtering condition (e.g., population > 5 million) directly inside a `facet_wrap()`."}
gapminder2007 %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp, colour = continent)) +
  geom_point(shape = 1) +
  facet_wrap(~pop > 50000000)

```

On this plot, the facet `FALSE` includes countries with a population less than 50 million people, and the facet `TRUE` includes countries with a population greater than 50 million people.

The tilde (~) in R denotes dependency. It is mostly used by statistical models to define dependent and explanatory variables, you will see it a lot in the second part of this book.

**(6)** Grey to white background - changing the theme

Overall, we can customise every single thing on a ggplot.
Font type, colour, size or thickness or any lines or numbers, background, you name it.
But a very quick way to change the appearance of a ggplot is to apply a different theme.
The signature ggplot theme has a light grey background and white grid lines (Figure \@ref(fig:chap04-fig-themes)). 

```{r chap04-fig-themes, echo = FALSE, fig.width=5, fig.height=3.5, fig.cap = "Some of the built-in ggplot themes (1) default (2) `theme_bw()`, (3) `theme_dark()`, (4) `theme_classic()`."}
themeplot = gapminder %>% 
  ggplot(aes(x = year, y = lifeExp)) +
  facet_wrap(~"LABEL")

(themeplot + (themeplot+theme_bw()))/((themeplot + theme_dark()) + (themeplot+theme_classic())) + plot_annotation(tag_levels = "1", tag_prefix = "(", tag_suffix = ")")

```


```{r fig.height=3.5, fig.width=7, fig.keep = 'none'}

step6 = gapminder2007 %>% 
  ggplot(aes(x = gdpPercap/1000, y = lifeExp, colour = continent)) +
  geom_point(shape = 1) +
  facet_wrap(~continent) +
  theme_bw()

```

As a final step, we are adding `theme_bw()` ("background white") to give the plot a different look. 
We have also divided the gdpPercap by 1000 (making the units "thousands of dollars per capita").
Note that you can apply calculations directly on ggplot variables (so how we've done `x = gdpPercap/1000` here).

This creates the last plot in Figure \@ref(fig:chap04-fig-steps).

This is how `ggplot()` works - by adding or modifying things one by one.

\clearpage

## Scatter plots/bubble plots

The ggplot anatomy (Section \@ref(chap04-gganatomy)) covered both scatter and strip plots (both created with `geom_point()`).
Another cool thing about this geom is that adding a size aesthetic makes it into a bubble plot.
For example, let's size the points by population. 

As you would expect from a "grammar of graphics plot", this is as simple as adding `size = pop` as an aesthetic:

```{r, fig.keep = 'none'}
gapminder2007 %>% 
  ggplot(aes(x = gdpPercap/1000, y = lifeExp, size = pop)) +
  geom_point()
```

With increased bubble sizes, there is some overplotting, so let's make the points hollow (`shape = 1`) and slightly transparent (`alpha = 0.5`):

```{r, fig.keep = 'none'}
gapminder2007 %>% 
  ggplot(aes(x = gdpPercap/1000, y = lifeExp, size = pop)) +
  geom_point(shape = 1, alpha = 0.5)
```

The resulting bubble plots are shown in Figure \@ref(fig:chap04-fig-bubble).


```{r chap04-fig-bubble, echo = FALSE, fig.width=0.6*10, fig.height = 0.6*4, fig.cap = "Turn the scatter plot from Figure \\@ref(fig:chap04-fig-steps):(2) to a bubble plot by (1) adding `size = pop` inside the `aes()`, (2) make the points hollow and transparent."}
p1 = gapminder2007 %>% 
  ggplot(aes(x = gdpPercap/1000, y = lifeExp, size = pop)) +
  geom_point()

p2 = gapminder2007 %>% 
  ggplot(aes(x = gdpPercap/1000, y = lifeExp, size = pop)) +
  geom_point(shape = 1, alpha = 0.5) +
  theme(legend.position = "none")

p1 + p2 + plot_annotation(tag_levels = "1", tag_prefix = "(", tag_suffix = ")")
```

Alpha is an aesthetic to make geoms transparent, its values can range from 0 (invisible) to 1 (solid).

## Line plots/time series plots

Let's plot the life expectancy in the United Kingdom over time:

```{r, fig.width = 4, fig.height = 1.5}
gapminder %>% 
  filter(country == "United Kingdom") %>% 
  ggplot(aes(x = year, y = lifeExp)) +
  geom_line()
```

As a recap, the steps in the code above are:

* Using the `gapminder` data send into a `filter()`
* inside the `filter()`, our condition is `country == "United Kingdom"`
* We initialise `ggplot()` and define our main variables: `aes(x = year, y = lifeExp)`
* we are using a new geom - `geom_point()`.

This is identical to how we used `geom_point()`.
In fact, by just changing `line` to `point` in the code above works - and instead of a continuous line you'll get a point at every 5 years as in the dataset.

But what if we want to draw multiple lines, e.g., for each country in the dataset. 
Let's send the whole dataset to `ggplot()` and `geom_line()`:

```{r, fig.keep = 'none'}
gapminder %>% 
  ggplot(aes(x = year, y = lifeExp)) +
  geom_line()
```

The reason you see this weird zig-zag in Figure \@ref(fig:chap04-fig-zigzag):1 is that, using the above code, `ggplot()` does not know which points to connect with which.
Yes, you know you want a line for each country, but you haven't told it that.
So for drawing multiple lines, we need to add a `group` aesthetic, in this case `group = country`:

```{r, fig.keep = 'none'}
mydata %>% 
  ggplot(aes(x = year, y = lifeExp, group = country)) +
  geom_line()
```

```{r chap04-fig-zigzag, echo = FALSE, fig.width = 7, fig.height=3, fig.cap = "The 'zig-zag plot' is a common mistake: using `geom_line()` (1) without a `group = ` aesthetic, (2) after adding `group = country`."}
p1 = gapminder %>% 
  ggplot(aes(x = year, y = lifeExp)) +
  geom_line()

p2 = mydata %>% 
  ggplot(aes(x = year, y = lifeExp, group = country)) +
  geom_line()

p1 + p2 + plot_annotation(tag_levels = "1", tag_prefix = "(", tag_suffix = ")")

```

This code works as expected (Figure \@ref(fig:chap04-fig-zigzag):2) - yes there is a lot of overplotting but that's just because we've included `r gapminder$country %>% n_distinct()` lines on a single plot.

### Exercise

Follow the step-by-step instructions to transform (Figure \@ref(fig:chap04-fig-zigzag):2) into this:


```{r, fig.width=0.8*10, echo = FALSE, fig.height=0.8*4}

gapminder %>% 
  ggplot(aes(x = year, y = lifeExp, group = country, colour=continent)) +
  geom_line() +
  facet_wrap(~continent) + 
  theme_bw() +
  scale_colour_brewer(palette = "Paired")

```

* Colour lines by continents: add `colour = continent` inside `aes()`
* Continents on separate facets: `+ facet_wrap(~continent)`
* Make the background white: `+ theme_bw()`
* Use a nicer colour scheme: `+ scale_colour_brewer(palette = "Paired")`

## Bar plots

There are two geoms for making bar plots - `geom_col()` and `geom_bar()`. In short: `geom_col()` plots values from your data directly (you define the `x` and `y` values), whereas `geom_bar()` will only take `x` values, the height of the bar (`y`) is the subgroups within `x` counted up. `geom_bar()` is basically a histogram for a categorical variable. 

**`geom_col()`**:

* requires two variables `aes(x = , y = )`
* `x` is categorical, `y` is continuous (numeric)

Let's plot the life expectancies in 2007 in these three countries:

```{r, fig.keep = 'none'}
gapminder2007 %>% 
  filter(country %in% c("United Kingdom", "France", "Germany")) %>% 
  ggplot(aes(x = country, y = lifeExp)) +
  geom_col() 
```

This gives us Figure \@ref(fig:chap04-fig-col):1.
We have also created another cheeky one using the same code but changing the scale of the y axis to be more dramatic (Figure \@ref(fig:chap04-fig-col):2).

```{r chap04-fig-col, echo = FALSE, fig.width=6.5, fig.cap = "Bar plots using `geom_col()`: (1) using the code example, (2) same plot but with `+ coord_cartesian(ylim=c(79, 81))` to manipulate the scale into something a lot more dramatic."}

p1 = gapminder2007 %>% 
  filter(country %in% c("United Kingdom", "France", "Germany")) %>% 
  ggplot(aes(x = country, y = lifeExp)) +
  geom_col() 

p2 = gapminder2007 %>% 
  filter(country %in% c("United Kingdom", "France", "Germany")) %>% 
  ggplot(aes(x = country, y = lifeExp)) +
  geom_col() +
  coord_cartesian(ylim=c(79, 81))

p1 + p2 + plot_annotation(tag_levels = "1", tag_prefix = "(", tag_suffix = ")")
```

\FloatBarrier


**`geom_bar()`**:

* requires a single variable `aes(x = )`
* this `x` should be a categorical variable
* `geom_bar()` then counts up the number of observations (rows) for this variable and plots them as bars.

Our `gapminder2007` tibble has a row for each country (see end of Section \@ref(chap04-data) to remind yourself). 
Therefore, if we use the `count()` function on the `continent` variable, we are counting up the number of countries on each continent (in this dataset^[The number of countries in this dataset is 142, whereas the United Nations have 193 member states]):

```{r}
gapminder2007 %>% 
  count(continent)
```


So `geom_bar()` basically runs the `count()` function and plots it (see how the bars on Figure \@ref(fig:chap04-fig-bar) are the same height as the values from `count(continent)`).

```{r chap04-fig-bar, fig.width=6.4, echo = FALSE, fig.cap = "`geom_bar()` counts up the number of observations for each group. (1) `gapminder2007 %>% ggplot(aes(x = continent)) + geom_bar()`, (2) same + a little bit of magic to reveal the underlying data."}
p1 = gapminder2007 %>% 
  ggplot(aes(x = continent)) +
  geom_bar()

p2 = gapminder2007 %>% 
  ggplot(aes(x = continent, colour = country)) +
  geom_bar(fill = NA) +
  theme(legend.position = "none")

p1 + p2 + plot_annotation(tag_levels = "1", tag_prefix = "(", tag_suffix = ")")

```


The first barplot in Figure \@ref(fig:chap04-fig-bar) is produced with just this:

```{r, fig.keep = 'none'}

gapminder2007 %>% 
  ggplot(aes(x = continent)) +
  geom_bar()

```

Whereas on the second one, we've asked `geom_bar()` to reveal the components (countries) in a colourful way:

```{r, fig.keep = 'none'}

gapminder2007 %>% 
  ggplot(aes(x = continent, colour = country)) +
  geom_bar(fill = NA) +
  theme(legend.position = "none")

```


We have added `theme(legend.position = "none")` to remove the legend - it includes all 142 countries and is not very informative in this case.
We're only including the colours for a bit of fun.

We're also removing the fill by setting it to NA (`fill = NA`). 
Note how we defined `colour = country` inside the `aes()` (as it's a variable), but we put the fill inside `geom_bar()` as a constant.
This was explained in more detail in steps (3) and (4) in the ggplot anatomy Section (\@ref(chap04-gganatomy)).


### colour vs fill

Figure \@ref(fig:chap04-fig-bar) also reveals the difference between a colour and a fill.
Colour is the border around a geom, whereas fill is inside it.
Both can either be set based on a variable in your dataset (this means `colour = ` or `fill = ` needs to be inside the `aes()` function), or they could be set to a fixed colour.

R has an amazing knowledge about different colours.
In addition to knowing what is "white", "yellow", "red", "green" etc. (meaning we can simply do `geom_bar(fill = "green")`) it also knows what "aquamarine", "blanchedalmond", "coral", "deeppink", "lavender", "deepskyblue" look like (amongst many many others, search the internet for "R colours" for a full list).


We can also use HEX colour codes, for example, `geom_bar(fill = "#FF0099")` is a very pretty pink.

### Exercise

Create this barplot of life expectancies in European countries (year 2007):

```{r, fig.width = 4.5, echo = FALSE, fig.height=5}
gapminder %>% 
  filter(year == 2007) %>%
  filter(continent == "Europe") %>% 
  ggplot(aes(x = fct_reorder(country, lifeExp), y = lifeExp)) +
  geom_col(colour = "deepskyblue", fill = NA) +
  coord_flip() +
  theme_classic()
  
```

Hints:

* If `geom_bar()` doesn't work try `geom_col()` or vice versa.
* `coord_flip()` to make the bars horizontal (it flips the `x` and `y` axes).
* `x = country` gets the country bars plotted in the alphabetical order, use `x = fct_reorder(country, lifeExp)` still inside the `aes()` to order the bars by their `lifeExp` values. Or try one of the other variables. (`pop`, `gdpPercap`) as the second argument to `fct_reorder()`.
* when using `fill = NA` also need to include a colour, we're using `colour = "deepskyblue"` inside the `geom_col()`.


## Box plots

Plot the distribution of life expectancies within each continent at year 2007:

* `filter(year == 2007) %>%`
* `x = continent, y = lifeExp`
* `+ geom_boxplot()`

```{r, fig.width=2.77, fig.height=2.75}

gapminder2007 %>% 
  ggplot(aes(x = continent, y = lifeExp)) +
  geom_boxplot()

```


### Exercise

Add individual (country) points on top of the box plot:

```{r, echo=F, fig.width=5, fig.height=4}

# optinal labels:
label_data = mydata %>% 
  filter(year == 2007) %>% 
  group_by(continent) %>% 
  filter(lifeExp == max(lifeExp) )


mydata %>% 
  filter(year == 2007) %>% 
  ggplot(aes(x = continent, y = lifeExp)) +
  geom_boxplot(outlier.shape = NA) +
  geom_label(data = label_data, aes(label=country), vjust = 0, size = 3) +
  geom_jitter(aes(colour=continent), width=0.3, alpha=0.8) + #width defaults to 0.8 of box width
  theme_bw()


```

Hint: Use `geom_jitter()` instead of `geom_point()` to reduce overlap by spreading the points horizontally. Include the `width=0.3` option to reduce the width of the jitter.

**Optional:**

Include text labels for the highest life expectancy country of each continent.

**Hint 1** Create a separate dataframe called `label_data` with the maximum countries for each continent:
```{r}
label_data = mydata %>% 
  filter(year == max(year)) %>% # same as year == 2007
  group_by(continent) %>% 
  filter(lifeExp == max(lifeExp) )
```

**Hint 2** Add `geom_label()` with appropriate `aes()`:
```{r, eval = FALSE}
+ geom_label(data = label_data, aes(label=country), vjust = 0)
```




## All other types of plots

These are just some of the main ones, see this gallery for more options: http://www.r-graph-gallery.com/portfolio/ggplot2-package/

And the `ggplot()` documentation: http://docs.ggplot2.org/

Remember that you can always combine different types of plots - i.e. add lines or points on bars, etc.


## Specifying `aes()` variables

The `aes()` variables wrapped inside `ggplot()` will be taken into account by all geoms. 
If you put `aes(colour = lifeExp)` inside `geom_point()`, only points will be coloured:

```{r, fig.width = 4, fig.height = 3}

mydata %>% 
  filter(continent == "Europe") %>% 
  ggplot(aes(x = year, y = lifeExp, group = country)) +
  geom_line() +
  geom_point(aes(colour = lifeExp))

```

## Extra: Optional exercises

### Exercise

Make this:


```{r, fig.height=10, fig.width=8}

mydata$dummy = 1  # create a column called "dummy" that includes number 1 for each country

mydata2007 = mydata %>% 
  filter(year==max(year)) %>% 
  group_by(continent) %>% 
  mutate(country_number = cumsum(dummy))  # create a column called "country_number" that
  # is a cumulative sum of the number of countries before it - basically indexing


mydata2007 %>% 
  ggplot(aes(x = continent)) +
  geom_bar(aes(colour=continent), fill = NA) +
  geom_text(aes(y = country_number, label=country), size=4, vjust=1, colour='black')+
  theme_void()

```

\newpage

### Exercise

Make this:

Hints: `coord_flip()`, `scale_color_gradient(...)`, `geom_segment(...)`, `annotate("text", ...)`

```{r, fig.width=7, fig.height=5}

mydata %>% 
  filter(continent == "Europe") %>% 
  ggplot(aes(y = fct_reorder(country, gdpPercap, .fun=max), x=lifeExp, colour=year)) +
  geom_point(shape = 15, size = 2) +
  theme_bw() +
  scale_colour_distiller(palette = "Greens", direction = 1) +
  geom_segment(aes(yend = "Switzerland", x = 85, y = "Bosnia and Herzegovina", xend = 85),
               colour = "black", size=1,
               arrow = arrow(length = unit(0.3, "cm"))) +
  annotate("text", y = "Greece", x=83, label = "Higher GDP per capita", angle = 90)

```




## Solutions

**4.2.1**

```{r, eval = FALSE, fig.width=6}

mydata %>% 
  filter(year == 2007) %>% 
  ggplot(  aes(x = gdpPercap/1000, #divide by 1000 to tidy the x-axis
               y = lifeExp,
               colour=continent,
               size=pop)) +
  geom_point(shape = 1) +
  facet_wrap(~continent) +
  theme_bw()

```

**4.3.1**

```{r, fig.width=10, eval = FALSE, fig.height=4}

mydata %>% 
  ggplot(  aes(x = year, y = lifeExp, group = country, colour=continent)) +
  geom_line() +
  facet_wrap(~continent) + 
  theme_bw() +
  scale_colour_brewer(palette = "Paired")

```

**which**

Add ` + 
geom_line(data = filter(mydata, country == "France"), colour = "blue")`

**4.4.1**

```{r, eval=F, fig.width=5, fig.height=4}

mydata %>% 
  filter(year == 2007) %>% 
  ggplot(aes(x = continent, y = lifeExp)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(aes(colour=continent), width=0.3, alpha=0.8) + #width defaults to 0.8 of box width
  theme_bw()


```


```{r, eval = FALSE}

mydata %>% 
  filter(year == 2007) %>% 
  ggplot(aes(x = continent, y = lifeExp)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(aes(colour=continent), width=0.3, alpha=0.8)
  theme_bw()


```

**4.5.1**

```{r, fig.width = 4, eval = FALSE}
mydata %>% 
  filter(year == 2007) %>%
  filter(continent == "Europe") %>% 
  ggplot(aes(x = country, y = lifeExp)) +
  geom_col(colour = "#91bfdb", fill = NA) +
  coord_flip() +
  theme_classic()
  
```

