# (PART) Medical Statistics {-}

# Tests for continuous outcome variables

## Load data

This session we will be using the gapminder dataset as in Session 4.

```{r, message=F}
library(tidyverse) 
library(gapminder)
library(broom)

mydata = gapminder
```

The first step of choosing the right statistical test is determining the type of variable you have.

Lets first have a look at some of our avalible data:

```{r, message=F}
mydata$continent %>% unique() # categorical

mydata$year %>% unique() # categorical

mydata$lifeExp %>% head() # continuous
```

## T-test
A t-test is used to compare the means of two groups of continuous variables.

### Plotting
Before you perform any statistical tests, you should always plot your data first to determine whether these have a "normal" distribution.
 
 - Histograms should form a symmetrical "bell-shaped curve".

 - Q-Q plots should fall along the 45 degree line.
 
 - Box-plots should be symmetrical and have few outliers.

### Histogram for each continent
```{r, fig.width=7}
theme_set(theme_bw())

mydata %>% 
	filter(year %in% c(2002, 2007)) %>%
	ggplot(aes(x = lifeExp)) +
	geom_histogram(bins = 10) +
	facet_grid(year~continent, scales = "free")
```

### Q-Q plot for each continent

With `ggplot()`, we can drawa Q-Q plot for each subgroup very efficiently:

```{r, fig.width=8, fig.height=3.5}
mydata %>% 
  filter(year %in% c(2002, 2007)) %>% 
  ggplot(aes(sample = lifeExp)) + 
  geom_point(stat = "qq") +
  facet_grid(year~continent)

```

Or we could save a subset of the data (e.g., "Americas" and year 2007 only) into a new variable (`subdata`) and use base R to draw a single Q-Q plot with less code:

```{r}

mydata %>% 
  filter(year == 2007) %>% 
  filter(continent == "Americas") -> subdata

qqnorm(subdata$lifeExp)
qqline(subdata$lifeExp)

```

### Boxplot of 2 years
```{r, fig.width=3, fig.height=3}
mydata %>% 
	filter(year %in% c(2002, 2007)) %>% 
	ggplot(aes(x = factor(year), y=lifeExp)) + #show that x = year errors: 
	geom_boxplot()                           #needs to be factor(year) or group=year
```

### Exercise
Make a histogram, Q-Q plot, and a box-plot for the life expectancy for a continent (you choose!) but for all years - does the data appear normally distributed?

## Two-sample t-tests

Lets perform a t-test on the "Americas" data as it appears normally distributed. 
We are savings the results of our t.test into a variable called t.result, but you can call it whatever you like (e.g. `myttest`).


```{r, fig.width=4, fig.height=4}
mydata %>% 
  filter(year %in% c(2002, 2007)) %>%
  filter(continent == "Americas") -> test.data

t.test(lifeExp~year, data=test.data)

mydata %>% 
  filter(year %in% c(2002, 2007)) %>%
  filter(continent == "Americas") %>% 
  t.test(lifeExp~year, data = .) -> t.result

t.result
```

### T-Test output
However, that output isn't in a useful format, let's investigate the output of the function `t.test()`. 
```{r}
names(t.result)
str(t.result) # or click on the blue button in the Environment tab
```

The structure of R's `t.test()` result looks a bit overwhelming. 
Fortunately, the `tidy()` function from `library(broom)` puts it into a neat dataframe for us:

```{r}

t.result <- tidy(t.result) # broom package puts it all in a data frame

```

try clicking on it in the Environment tab now.

Thus, now we understand the output structure we can extract any result.
```{r}
t.result$p.value
```


### Exercise

1. Select any 2 years in any continent and perform a t-test to determine whether the life expectancy is signficantly different. 
Remember to plot your data first.

2. Extract only the p-value from your `t.test()` output.


## One sample t-tests
However, we don't always want to compare 2 groups or sometimes we don't have the data to be able to.

Let's investigate whether the mean life expectancy in each continent significant different to 77 years in 2007.

```{r}

mydata %>% 
  filter(year==2007, continent=='Europe') -> subdata

# Standard one-sample t-test
t.test(subdata$lifeExp, mu=77)

```

### Exercise
1. Select a different year, different continent, and different age to compare the mean life expectancy to.

2. Replace mu=77 with mu=0 (the default value). How does this affect your result?


## ANOVA
In some cases, we may also want to test more than two groups to see if they are signficantly different.

### Plotting
For example, lets plot the life expectancy in 2007 accross 3 continents.
```{r, fig.width=3, fig.height=3}
mydata %>% 
	filter(year == 2007) %>% 
	filter(continent %in% c("Americas", "Europe", "Asia")) %>% 
	ggplot(aes(x = continent, y=lifeExp)) +
	geom_boxplot()
```

### Analysis
ANOVA tests are useful for testing for the presence of signficant differences between more than two groups or variables.
```{r}
mydata %>% 
  filter(year == 2007) %>% 
  filter(continent %in% c("Americas", "Europe", "Asia")) -> subdata

fit = aov(lifeExp~continent, data = subdata) 

summary(fit)

mydata %>% 
  filter(year == 2007) %>% 
  filter(continent %in% c("Americas", "Europe", "Asia")) %>% 
  aov(lifeExp~continent, data = .) %>% 
	tidy()
```

### Check assumptions
```{r, fig.width=5, fig.height=5}
# Check ANOVA assumptions 
par(mfrow=c(2, 2)) # 4 plots in 2 x 2 grid
plot(fit)
```

### Perform pairwise tests
The ANOVA test was significant, indicating that there is a signficant difference in the mean life expectancy across those continents.

But which continents are significantly different, and can we quantify this difference as a p-value?

```{r}
mydata %>% 
  filter(year == 2007) %>% 
  filter(continent %in% c("Americas", "Europe", "Asia")) -> subdata

pairwise.t.test(subdata$lifeExp, subdata$continent)

# or equivalently, without saving the subset in a separate variable:
# sending it into the test using pipes only
mydata %>% 
  filter(year == 2007) %>% 
  filter(continent %in% c("Americas", "Europe", "Asia")) %>% 
  pairwise.t.test(.$lifeExp, .$continent, data=.) %>% 
  tidy()

```

F1 for help to see options for `pairwise.t.test()`.

### Top tip: the cut() function

A great way of easily converting a continuous variable to a categorical variable is to use the cut() function
```{r}

pop_quantiles = quantile(mydata$pop)

mydata %>% 
	mutate(pop.factor = cut(pop, breaks=pop_quantiles)) -> mydata
```

### Exercise

When we used `cut()` to divide country populations into quantiles, the labels it assigned are not very neat:

```{r}

mydata$pop.factor %>% levels()

```

Use `fct_recode()` to change them to something nicer, e.g., "Tiny", "Small", "Medium", "Large":

```{r}
mydata$pop.factor %>% 
  fct_recode("Tiny"   = "(6e+04,2.79e+06]",
             "Small"  = "(2.79e+06,7.02e+06]",
             "Medium" = "(7.02e+06,1.96e+07]",
             "Large"  = "(1.96e+07,1.32e+09]") -> mydata$pop.factor 
```


### Exercise
Perform ANOVA to test for a difference in mean life expectancy by country population factor (`mydata$pop.factor`). Remember to plot data first

```{r, fig.width=4, fig.height=3}
mydata %>% 
	filter(year == 2007) %>% 
	ggplot(aes(x=pop.factor, y=lifeExp))+
	geom_boxplot()

mydata %>% 
	filter(year == 2007) %>% 
	aov(.$lifeExp ~ .$pop.factor, data=.) %>% 
	summary()
```



## Non-parametric data
If your data is not parametric (e.g. not normally distributed) then the usual t-test is invalid. In this case there are 2 options:

1. Non-parametric statistical tests.

2. "Transform" the data to fit a normal distribution (*not covered here*) so that a t-test can be used.

### Plotting
Lets plot the life expectancy within Africa in 1997, 2002, and 2007.
```{r, fig.width=3, fig.height=3}
# African data is not normally distributed
mydata %>% 
  filter(year %in% c(1997, 2002, 2007)) %>%
  filter(continent == "Africa") %>% 
  ggplot(aes(x = lifeExp)) +
  geom_histogram(bins = 10, fill=NA, colour='black') +
  facet_grid(year~continent)

mydata %>% 
  filter(year %in% c(1997, 2002, 2007)) %>%
  filter(continent == "Africa") %>% 
  group_by(year) %>% 
  summarise(avg = mean(lifeExp), med = median(lifeExp))
```


### Exercise: Non-parametric testing
Mann-Whitney U test is also called the Wilcoxon rank sum test (note the Wilcoxon signed rank test is for paried data).

Is there a significant increase in the life expectencies for African countries between 1992 and 2007? How about 1982 and 2007?

```{r, fig.width=3, fig.height=2}
mydata$year %>%  unique()

mydata %>% 
  filter(continent == "Africa") %>% 
  group_by(year) %>% 
  summarise(mean = mean(lifeExp), median = median(lifeExp)) %>% 
  ggplot(aes(x = year, y = median)) +
	  geom_line()

```


```{r, fig.width=4, fig.height=3}
mydata %>% 
  filter(continent == "Africa") %>% 
  ggplot(aes(x = factor(year), y=lifeExp)) + #demonstrate that needs to be factor(year), not year
  geom_boxplot()

mydata %>% 
  filter(year %in% c(1992, 2007)) %>%
  filter(continent == "Africa") %>% 
  wilcox.test(lifeExp~year, data=.)
```

## Solutions

**5.2.2**

```{r, fig.width=4, fig.height=4, eval = FALSE}

mydata %>% 
  filter(continent == "Europe") %>% 
  ggplot(aes(x = lifeExp)) + 
  geom_histogram() +
  facet_wrap(~year)

mydata %>% 
  filter(continent == "Europe") %>% 
  ggplot(aes(sample = lifeExp)) + 
  geom_point(stat = "qq") +
  facet_wrap(~year)

mydata %>% 
  filter(continent == "Europe") %>% 
  ggplot(aes(y = lifeExp, x = factor(year))) + 
  geom_boxplot()
  
```




## Advanced example
This is a complex but useful example which shows you the power of the syntax. Here multiple t-tests are performed and reported with just a few lines of code. 

Performing t-tests across all continents at once:

```{r, fig.width=6, fig.height=4}
mydata %>% 
  filter(year %in% c(1997, 2007)) %>% 
  group_by(continent) %>%
	do(
		tidy(
			t.test(lifeExp~year, data=.)
		)
	)
```
