# (PART) Data analysis {-}

```{r include=FALSE}
library(ggplot2)
theme_set(theme_bw())
```

In the second part of this book, we focus specifically on the business of data analysis.
That is, formulating clear questions and seeking to answer them using available datasets. 

Again, we emphasise the importance of understanding the underlying data through visualisation, rather than relying on statistical tests or, heaven forbid, the p-value alone. 

There are five chapters. 
Testing for continuous outcome variables (6) leads naturally into Linear regression (7). 
We would expect the majority of actual analysis done by readers to be using the methods in chapter 7 rather than 6. 
Similarly, Testing for categorical outcome variables (8) leads naturally to Logistic regression (9), where we would expect the majority of work to focus. 
Chapters 6 and 8 however do provide helpful reminders of how to prepare data for these analyses and shouldn't be skipped.
Time-to-event data introduces survival analysis and includes sections on the manipulattion of dates. 

# Working with continuous outcome variables{#chap06-h1}
\index{continuous data@\textbf{continuous data}}

> Continuous data can be measured.  
> Categorical data can be counted.

## Continuous data

Continuous data is everywhere in healthcare. 
From physiological measures in patients such as systolic blood pressure or pulmonary function tests, through to populations measures like life expectancy or disease incidence, the analysis of continuous outcome measures is common and important.

Our goal in most health data questions, is to draw a conclusion on a comparison between groups. 
For instance, understanding differences in life expectancy between the year 2002 and 2007 is more useful than simply describing the average life expectancy across all of time. 

The basis for comparisons between continuous measures is the *distribution* of the data. 
That word, as many which have a statistical flavour, brings on the sweats in a lot of people. 
It needn't. 
By distribution, we are simply referring to the shape of the data. 

## The Question

The examples in this chapter all use the data introduced previously from the amazing [Gapminder project](https://www.gapminder.org/). 
We will start by looking at the life expectancy of populations over time and in different geographical regions. 

## Get the data

```{r, message=F}
# Load packages
library(tidyverse)
library(finalfit)
library(gapminder)

# Create object mydata from object gapminder
mydata = gapminder
```

## Check the data{#chap06-h2-check}

It is vital that data is carefully inspected when first read (for help reading data into R see \@ref(chap02-h2-reading-data-into-r)).
The three functions below provide a clear summary allowing errors or miscoding to be quickly identified. 
It is particularity important to ensure that any missing data is identified (see Chapter \@ref(chap13-h1). 
If you don't do this you will regret it! 
There are many times when an analysis has got to a relatively advanced stage before research realised the dataset was incomplete. 

```{r}
glimpse(mydata) # each variable as line, variable type, first values
missing_glimpse(mydata) # missing data for each variable
```

```{r eval = FALSE}
ff_glimpse(mydata) # summary statistics for each variable
```

\index{functions@\textbf{functions}!glimpse}
\index{functions@\textbf{functions}!missing\_glimpse}
\index{functions@\textbf{functions}!ff\_glimpse}

```{r include=FALSE}
sum_mydata = ff_glimpse(mydata) # summary statistics for each variable
```

```{r message=FALSE, echo=FALSE}
sum_mydata = ff_glimpse(mydata) # summary statistics for each variable
library(knitr)
library(kableExtra)
sum_mydata[[1]] %>%
  select(-c(5, 8, 9, 11, 12)) %>% 
  kable(row.names = FALSE, align = c("l", "l", "l", "r", "r", "r", "r", "r", "r", "r"), 
        booktabs = TRUE, caption = "Gapminder dataset, ff\\_glimpse: continuous", 
        linesep = c("", "", "\\addlinespace")) %>%
  kable_styling(latex_options = c("scale_down", "hold_position")) %>% 
  column_spec(1, width = "4cm")
sum_mydata[[2]] %>% 
  select(-c(5, 9)) %>% 
  kable(row.names = FALSE, align = c("l", "l", "l", "r", "r", "r", "r", "r", "r", "r"), 
        booktabs = TRUE, caption = "Gapminder dataset, ff\\_glimpse: categorical", 
        linesep = c("", "", "\\addlinespace")) %>%
  kable_styling(latex_options = c("scale_down", "hold_position")) %>%  
  column_spec(6, width = "3cm") %>% 
  column_spec(7, width = "3cm")
```



As can be seen, there are 6 variables, 4 are continuous and 2 are categorical. The categorical variables are already identified as `factors`. There are no missing data. 

## Plot the data

We will start by comparing life expectancy between the 5 continents of the world in two different years. 
Always plot your data first. 
Never skip this step!
We are particularly interested in the distribution. 
There's that word again. 
The shape of the data. 
Is it normal? 
Is it skewed? 
Does it differ between regions and years?

There are three useful plots which can help here: 

- Histograms: examine shape of data and compare groups;
- Q-Q plots: are data normally distributed?
- Box-plots: identify outliers, compare shape and groups.

### Histogram

```{r chap06-fig-hist-life-year, fig.width=7, fig.height=3.5, fig.cap="Histogram: country life expectancy by continent and year"}
mydata %>% 
  filter(year %in% c(2002, 2007)) %>%
  ggplot(aes(x = lifeExp)) +       # remember aes()
  geom_histogram(bins = 20) +      # histogram with 20 bars
  facet_grid(year ~ continent)     # add scale="free" for axes to vary                                 
```

What can we see? 
That life expectancy in Africa is lower than in other regions. 
That we have little data for Oceania given there are only two countries included, Australia and New Zealand. 
That Africa and Asia have great variability in life expectancy by country than in the Americas or Europe. 
That the data follow a reasonably normal shape, with Africa 2002 a little right skewed. 

### Q-Q plot{#chap06-h3-qq-plot}

A quantile-quantile sounds complicated but is not. 
It is simply a graphical method for comparing the distribution (think shape) of our own data to a theoretical distribution, such as the normal distribution. 
In this context, quantiles are just cut points which divide our data into bins each containing the same number of observations.
For example, if we have the life expectancy for 100 countries, then quartiles (note the quar-) for life expectancy are the three ages which split the observations into 4 groups each containing 25 countries. 
A Q-Q plot simply plots the quantiles for our data against the theoretical quantiles for a particular distributions (the default shown below is the normal distribution). 
If our data follow that distribution (e.g. normal), then we get a 45 degree line on the plot. 

```{r chap06-fig-qq-life-year, fig.width=7, fig.height=3.5, fig.cap="Q-Q plot: country life expectancy by continent and year"}
mydata %>% 
  filter(year %in% c(2002, 2007)) %>%
  ggplot(aes(sample = lifeExp)) +      # Q-Q plot requires `sample
  geom_qq() +                          # defaults to normal distribution
  geom_qq_line() +                     # add 45 degree line
  facet_grid(year ~ continent)
```

\index{plotting@\textbf{plotting}!geom\_qq}
\index{plotting@\textbf{plotting}!geom\_qq\_line}

What can we see?
We are looking to see if the data follow the 45 degree line which is included in the plot.
These do reasonably, except for Africa which is curved upwards at each end, suggesting a skew. 

We are frequently asked about performing a hypothesis test to check the assumption of normality, such as the Shapiro-Wilk normality test. 
We do not recommend this, simply because it is often non-significant when the number of observations is small but the data look skewed, and often significant when the number of observations is high but the data look reasonably normal on inspection of plots. 
It is therefore not useful in practice - common sense should prevail.

### Boxplot
\index{plotting@\textbf{plotting}!geom\_boxplot}
\index{plotting@\textbf{plotting}!boxplot}

Boxplots are our preferred method for comparing a continuous variable such as life expectancy with a categorical explanatory variable. 
It is much better than a bar plot, or a bar plot with error bars, sometimes called a dynamite plot. 

The box represents the median and interquartile range (where 50% of the data sits). 
The lines (whiskers) by default are 1.5 times the interquartile range.
Outliers are represented as points. 

Thus it contains information, not only on central tenancy (median), but on the variation in the data and the distribution of the data, for instance a skew should be obvious. 

```{r chap06-fig-boxplot-life-year, fig.width=7, fig.height=3, fig.cap="Boxplot: country life expectancy by continent and year"}
mydata %>% 
  filter(year %in% c(2002, 2007)) %>%
  ggplot(aes(x = continent, y = lifeExp)) +
  geom_boxplot() +
  facet_grid(. ~ year)                    # spread by year, note `.`
```

What can we see?
The median life expectancy is lower in Africa than in any other continent. 
The variation in life expectancy is greatest in Africa and smallest in Oceania. 
The data in Africa looks skewed, particularly in 2002 - the lines/whiskers are unequal lengths. 

We can add further arguments 

```{r chap06-fig-boxplot-jitter, fig.width=7, fig.height=3, fig.cap="Boxplot with jitter points: country life expectancy by continent and year"}
mydata %>%  
  filter(year %in% c(2002, 2007)) %>%
  ggplot(aes(x = factor(year), y = lifeExp)) +
  geom_boxplot(aes(fill = continent)) +     # add colour to boxplots
  geom_jitter(alpha = 0.4) +                # alpha = transparency
  facet_grid(. ~ continent) +               # spread by year, note `.`
  theme(legend.position = "none") +         # remove legend
  xlab("Year") +                            # label x-axis
  ylab("Life expectancy (years)") +         # label y-axis
  ggtitle(
    "Life expectancy by continent in 2002 v 2007") # add title
```

## Compare the means of two groups

### T-test
\index{t-test@\textbf{t-test}}
A *t*-test is used to compare the means of two groups of continuous variables. Volumes have been written about this else where, and we won't rehearse it here. 

There are various variations on the *t*-test.
We will use two here. 
The most useful in our context is a two-sample test of independent groups. 
Repeated-measures data, such as comparing the same countries between years, can be analysed using a paired *t*-test.

<!-- Any more here? -->


### Two-sample *t*-tests
\index{t-test@\textbf{t-test}!two-sample}

Referring to the first figure, let's compare life expectancy between Asia and Europe for 2007. 
What is imperative, is that you decide what sort of difference exists by looking at the boxplot, rather than relying on the *t*-test output. 
The median for Europe is clearly higher than in Asia. 
The distributions overlap, but it looks likely that Europe has a higher life expectancy than Asia. 

```{r}
ttest_data = mydata %>%                      # save as object testdata
  filter(year == 2007) %>%                   # 2007 only
  filter(continent %in% c("Asia", "Europe")) # Asia/Europe only

ttest_result = 
  t.test(lifeExp ~ continent, data = ttest_data) # Base R t.test
ttest_result
```

\index{functions@\textbf{functions}!t.test}

The Welch two-sample t-test is the most flexible and copes with differences in variance (variability) between groups, as in this example. 
The difference in means is provided at the bottom of the output. 
The *t*-value, degrees of freedom (df) and p-value are all provided. 
The p-value is 0.00003. 

The base R output is not that easy to utilise. 
For reference, the results can be explored and exported. 
However, more straightforward methods are provided below. 

```{r}
names(ttest_result)  # Names of elements of result object
str(ttest_result)    # Details of result object
ttest_result$p.value # Extracted element of result object
```

The `broom` package provides useful methods for 'tidying' common model outputs into a `tibble`.  
The whole analysis can be constructed as a single piped function. 

```{r}
library(broom)
mydata %>%
  filter(year == 2007) %>%                        # 2007 only
  filter(continent %in% c("Asia", "Europe")) %>%  # Asia/Europe only
  t.test(lifeExp ~ continent, data = .) %>%  
  tidy()

```

### When pipe sends data to the wrong place: use `, data = .` to direct it

In the code above, the `, data = .` bit is necessary because the pipe usually sends data to the beginning of function brackets. So `mydata %>% t.test(lifeExp ~ continent)` would be equivalent to `t.test(mydata, lifeExp ~ continent`. 
However, this is not an order that `t.test()` will accept. 
`t.test()` wants us to specify the formula first, and then wants the data these variables are present in. 
So we have to use the `.` to tell the pipe to send the data to the second argument of `t.test()`, not the first.

### Paired *t*-tests
\index{t-test@\textbf{t-test}!paired}

Consider that we want to compare the difference in life expectancy in Asian countries between 2002 and 2007.
The overall difference is not impressive in the boxplot. 

We can plot differences at the country level directly. 

```{r chap06-fig-line-life-asia, fig.height=3, fig.width=3, fig.cap="Line plot: Change in life expectancy in Asian countries from 2002 to 2007"}
paired_data = mydata %>%               # save as object paired_data
  filter(year %in% c(2002, 2007)) %>%  # 2002 and 2007 only
  filter(continent == "Asia")          # Asia only

paired_data %>%      
  ggplot(aes(x = year, y = lifeExp, 
             group = country)) +       # for individual country lines
  geom_line()
```

\index{plotting@\textbf{plotting}!geom\_line}

What is the difference in life expectancy for each individual country?
We don't usually have to produce this directly, but here is one method. 

```{r fig.height=3, fig.width=3}
paired_table = paired_data %>%        # save object paired_data
  select(country, year, lifeExp) %>%  # select vars interest
  spread (year, lifeExp) %>%          # make wide table
  mutate(
    dlifeExp = `2007` - `2002`        # difference in means
  )

paired_table

# Mean of difference in years
paired_table %>% summarise( mean(dlifeExp) )

```

On average, therefore, there is an increase in life expectancy of 1.5 years in Asian countries between 2002 and 2007. 
Let's test whether this number differs from zero with a paired *t*-test. 

```{r}
paired_data %>% 
  t.test(lifeExp ~ year, data = .) # Include paired = TRUE
```

\index{functions@\textbf{functions}!t.test}

The results show a highly significant difference. 
As an exercise you can repeat this analysis simply comparing the means in an unpaired manner. 
The resulting p-value is `R paired_data %>% t.test(lifeExp ~ year, data = .)$p.value`.

Why is there such a difference between the two approaches?
This emphasises just how important it is to plot the data first. 
The average difference of 1.5 years is highly consistent between countries, as shown on the line plot, and this differs from zero.
It is up to you the investigator to interpret the relevance of the effect size of 1.5 y in reporting the finding.

## Compare the mean of one group

### One sample *t*-tests
\index{t-test@\textbf{t-test}!one-sample}

We can use a *t*-test to determine whether the mean of a distribution is different to a specific value. 

The paired *t*-test above is is equivalent to a one-sample *t*-test on the calculated difference in life expectancy being different to zero.

```{r}
t.test(paired_table$dlifeExp)
```

We can compare to values other than zero. 
For instance, we can test whether the mean life expectancy in each continent was significantly different to 77 years in 2007.
We have included some extra code here to demonstrate how to run multiple base R tests in one pipe function. 

```{r}
mydata %>% 
  filter(year == 2007) %>%          # 2007 only
  group_by(continent) %>%           # split by continent
  do(                               # dplyr function
    t.test(.$lifeExp, mu = 77) %>%  # compare mean to 77 years 
      tidy()                        # tidy into tibble
  )
```

*group_modify option - decision on whether to switch to this needed*

```{r}
mydata %>% 
  filter(year == 2007) %>%           # 2007 only
  group_by(continent) %>%            # split by continent
  group_modify(                      # dplyr function
    ~ t.test(.$lifeExp, mu = 77) %>% # compare mean to 77 years 
      tidy()                         # tidy into tibble
  )
```

The mean life expectancy for Europe and Oceania do not differ from 77, while the others to to varying degrees. 
In particular, look at the confidence intervals of the tables and whether they include or exclude 77.


## Compare the means of more than two groups

It may be that our question is set around a hypothesis involving more than two groups. 
For example, we may be interested in comparing life expectancy across 3 continents such as the Americas, Europe and Asia.

### Plot the data

```{r, fig.width=3, fig.height=3, fig.cap="Boxplot: Life expectancy in selected continents for 2007"}
mydata %>% 
  filter(year == 2007) %>% 
  filter(continent %in% 
           c("Americas", "Europe", "Asia")) %>% 
  ggplot(aes(x = continent, y=lifeExp)) +
  geom_boxplot()
```

### ANOVA
\index{analysis of variance (ANOVA)}

Analysis of variance is a collection of statistical tests which can be used to test the difference in means between two or more groups. 

In base R form, it produces an ANOVA table which includes an F-test. 
This so-called omnibus test tells you whether there are any differences in the comparison of means of the included groups. 
Again, it is important to plot carefully and be clear what question you are asking.  

```{r}
aov_data = mydata %>% 
  filter(year == 2007) %>% 
  filter(continent %in% c("Americas", "Europe", "Asia"))

fit = aov(lifeExp ~ continent, data = aov_data) 
summary(fit)
```

\index{functions@\textbf{functions}!aov}

We can conclude from this, that there is a difference in the means between at least two pairs of the included continents. 
As above, the output can be neatened up using the `tidy` function. 

```{r}
library(broom)
mydata %>% 
  filter(year == 2007) %>% 
  filter(continent %in% c("Americas", "Europe", "Asia")) %>% 
  aov(lifeExp~continent, data = .) %>% 
  tidy()
```

### Assumptions

As with the normality assumption of the *t*-test, there are assumptions of the ANOVA model). 
These are covered in detail in the linear regression chapter and will not be repeated here. 
Suffice to say that diagnostic plots can be produced to check that the assumptions are fulfilled. 

```{r fig.height=6, fig.width=6, fig.cap="Diagnostic plots: ANOVA model of life expectancy by continent for 2007"}
par(mfrow=c(2,2))
plot(fit)
par(mfrow=c(1,1))
```

### Pairwise testing and multiple comparisons
\index{pairwise testing}

When the F-test is significant, we will often want to proceed to try and determine where the differences lie. 
This should of course be obvious from the boxplot you have made. 
However, some are fixated on the p-value!

```{r}
pairwise.t.test(aov_data$lifeExp, aov_data$continent, 
                p.adjust.method = "bonferroni")
```

\index{functions@\textbf{functions}!pairwise.t.test}

A matrix of pairwise p-values is produced.
Here we can that there is good evidence of a difference in means between Europe and Asia. 

The p-values are corrected for multiple comparisons. 
When performing a hypothesis test at the 5% level (alpha = 0.05), there is a 5% chance of a type 1 error. 
That is, a 1 in 20 chance of concluding a difference exists when it in fact does not (formally, this is a 1 in 20 chance of rejecting a true null hypothesis). 
As more simultaneous statistical tests are performed, the chance of a type 1 error increases. 

There are three approaches to this. The first, is not to perform any correction at all. 
Some advocate that the best approach is simply to present the results of all the tests that were performed, and let the sceptical reader make adjustments for themselves. 
This is attractive, but presupposes a sophisticated readership who will take the time to consider the results in their entirety. 

The second and classical approach, is to control for the so-called family-wise error rate. 
The "Bonferroni" correction is probably the most famous and most conservative, where the threshold for significance is lowered in proportion to the number of comparisons made. 
For example, if three comparisons are made, the threshold for significance is lowered to 0.017. 
Equivalently, any particular p-value can be multiples by 3 and the value compared to a threshold of 0.05, as is done above. 
The Bonferroni method is particular conservative, meaning that type 2 errors may occur (failure to identify true differences, or false negatives) in favour or minimising type 1 errors (false positives).
\index{Bonerroni correction}

The third newer approach controls false-discovery rate. 
The development of these methods has been driven in part by the needs of areas of science where many different statistical tests are performed at the same time, for instance, examining the influence of 1000 genes simultaneously. 
In these hypothesis-generating settings, a higher tolerance to type 1 errors may be preferable to missing potential findings through type 2 errors. 
\index{false discovery rate}
You can see in our example, that the p-values are lower with the `fdr` correction when compared to the `Bonferroni` correction. 

```{r}
pairwise.t.test(aov_data$lifeExp, aov_data$continent, 
                p.adjust.method = "fdr")
```

\index{functions@\textbf{functions}!pairwise.t.test}

Try not to get too hung up on this. 
Be sensible. 
Plot the data and look for differences. 
Focus on effect size. 
For instance, what is the actual difference in life expectancy in years, rather than the p-value of a comparison test. 
Choose a method which fits with your overall aims. 
If you are generating hypotheses which you will proceed to test with other methods, the `fdr` approach may be preferable.
If you are trying to capture robust effect and want to minimise type 2 errors, use a family-wise approach. 


## Non-parametric data
\index{non-parametric tests@\textbf{non-parametric tests}}

What if your data is different shape to normal or the ANOVA assumptions are not fulfilled (see linear regression chapter). 
As always, be sensible!
Would your data be expected to be normally distributed given the data-generating process?
For instance, if you examining length of hospital stay it is likely that your data are highly right skewed - most patients are discharged from hospital in a few days while a smaller number stay for a long time. 
Is a comparison of means ever going to be the correct approach here?
Perhaps you should consider a time-to-event analysis for instance (see chapter \@ref(ch10-h1)).

If a comparison of means approach is reasonable, but the normality assumption is not fulfilled there are two approaches, 

1. Transform the data;
2. Perform non-parametric tests.

### Transforming data
\index{tranformations}

Remember, the Welch *t*-test is reasonably robust to divergence from the normality assumption, so small deviations can be safely ignored. 

Otherwise, the data can be transformed to another scale to deal with a skew. 
A natural `log` scale is common. 

```{r echo=FALSE, message=FALSE}
library(kableExtra)
data.frame(
  Distribution = c("Moderate right skew (+)", "Substantial right skew (++)", "Substantial right skew (+++)"),
  Transformation = c("Square-root", "Natural log*", "Base-10 log*"),
  Function = c('sqrt()', 'log()', 'log10()')
) %>% 
  # mutate( 
  #   Function = kableExtra::cell_spec(Function, monospace = TRUE)) %>% # Not working
  knitr::kable(booktabs = TRUE,
               caption = "Transformations that can be applied to skewed data") %>% 
  footnote(general = "For left skewed data, subtract all values from a constant greater than the maximum value.",
           symbol = "If data contain zero values, add a small constant to all values.") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"))
```


```{r fig.height=3, fig.width=6, fig.cap="Histogram: Log transformation of life expectancy for countries in Africa 2002"}
africa_data = mydata %>%                # save as africa_data
  filter(year == 2002) %>%              # only 2002
  filter(continent == "Africa") %>%     # only Africa
  select(country, lifeExp) %>%          # only these variables
  mutate(
    lifeExp_log = log(lifeExp)          # log life expectancy
  )
head(africa_data)                       # inspect

africa_data %>% 
  gather(key, lifeExp, -country) %>%     # gather vals to same column
  ggplot(aes(x = lifeExp)) +             
  geom_histogram(bins = 15) +            # make histogram
  facet_grid(. ~ key, scales = "free")   # facet & axes free to vary
```

This has worked well here. 
The right skew on the Africa data has been dealt with by the transformation. 
A parametric test such as a *t*-test can now be performed. 

### Non-parametric test for comparing two groups
\index{non-parametric tests@\textbf{non-parametric tests}!Mann-Whitney U}
\index{non-parametric tests@\textbf{non-parametric tests}!Wilcoxon rank sum}

The Mann-Whitney U test is also called the Wilcoxon rank-sum test and uses a rank-based method to compare two groups (note the Wilcoxon signed-rank test is for paired data).
We can use if to test for a difference in life expectancies for African countries between 1982 and 2007.
Let's do a histogram, Q-Q plot and boxplot first. 

```{r fig.height=4, fig.width=6, fig.cap="Panels plots: histogram, Q-Q, boxplot for life expectancy in Africa 1992 v 2007"}
africa_plot = mydata %>%                          
  filter(year %in% c(1982, 2007)) %>%      # only 1982 and 2007
  filter(continent %in% c("Africa"))       # only Africa

p1 = africa_plot %>%                       # save plot as p1
  ggplot(aes(x = lifeExp)) + 
  geom_histogram(bins = 15) +
  facet_grid(. ~ year)

p2 = africa_plot %>%                       # save plot as p2
  ggplot(aes(sample = lifeExp)) +          # `sample` for Q-Q plot
  geom_qq() + 
  geom_qq_line() + 
  facet_grid(. ~ year)

p3 = africa_plot %>%                       # save plot as p3
  ggplot(aes(x = factor(year),
             y = lifeExp)) +               # change year to factor
  geom_boxplot(aes(fill = factor(year))) + # colour boxplot
  geom_jitter(alpha = 0.4) +               # add data points
  theme(legend.position = "none")          # remove legend

library(patchwork)                         # great for combining plots
p1 / p2 | p3

```

\index{plotting@\textbf{plotting}!patchwork}

The data is a little skewed based on the histograms and Q-Q plots. 
The difference between 1982 and 2007 is not particularly striking on the boxplot.

```{r}
africa_plot %>% 
  wilcox.test(lifeExp ~ year, data = .)
```

\index{functions@\textbf{functions}!wilcox.test}

### Non-parametric test for comparing more than two groups
\index{non-parametric tests@\textbf{non-parametric tests}!!Kruskal-Wallis}

The non-parametric equivalent to ANOVA, is the Kruskal-Wallis test. 
It can be used in base R, or via the finalfit package below. 

```{r}
library(broom)
mydata %>% 
  filter(year == 2007) %>% 
  filter(continent %in% c("Americas", "Europe", "Asia")) %>% 
  kruskal.test(lifeExp~continent, data = .) %>% 
  tidy()
```

\index{functions@\textbf{functions}!kruskal.test}

## Finalfit approach

The finalfit package provides an easy to use interface for performing non-parametric hypothesis tests.
Any number of explanatory variables can be tested against a so-called dependent variable. 
In this case, this is equivalent to a typical Table 1 in healthcare study. 

```{r eval=FALSE}
dependent = "year"
explanatory = c("lifeExp", "pop", "gdpPercap")
mydata %>%                          
  filter(year %in% c(1982, 2007)) %>%       # only 1982 and 2007
  filter(continent == "Africa") %>%         # only Africa
  mutate(
    year = factor(year)                     # change year to factor
  ) %>% 
  summary_factorlist(dependent, explanatory,
                     cont = "median", p = TRUE)
```

```{r echo=FALSE}
dependent = "year"
explanatory = c("lifeExp", "pop", "gdpPercap")
mydata %>%                          
  filter(year %in% c(1982, 2007)) %>%       # only 1982 and 2007
  filter(continent == "Africa") %>%         # only Africa
  mutate(
    year = factor(year)                     # change year to factor
  ) %>% 
  summary_factorlist(dependent, explanatory,
                     cont = "median", p = TRUE) %>% 
  knitr::kable(row.names = FALSE, booktabs = TRUE, 
               align = c("l", "l", "r", "r", "r", "r"),
               caption = "Life expectancy, population and GDPperCap in Africa 1982 v 2007") %>% 
    kable_styling(latex_options = c("scale_down", "hold_position"))
```

\index{functions@\textbf{functions}!summary\_factorlist}

## Conclusions

Continuous data is frequently encountered in a healthcare setting. 
Liberal use of plotting is required to really understand the underlying data. 
Comparisons can easily made between two or more groups of data, but always remember what you are actually trying to analyse and don't become fixated on the p-value. 
In the next chapter, we will explore the comparison of two continuous variables together with multivariable models of datasets.   

## Exercises

### Exercise 1

Make a histogram, Q-Q plot, and a box-plot for the life expectancy for a continent of your choice, but for all years. 
Do the data appear normally distributed?

### Exercise 2

1. Select any 2 years in any continent and perform a *t*-test to determine whether mean life expectancy is significantly different. 
Remember to plot your data first.

2. Extract only the p-value from your `t.test()` output.

### Exercise 3

In 2007, in which continents did mean life expectancy differ from 70.

### Exercise 4

1. Use ANOVA to determine if the population changed significantly through the 1990s/2000s in individual continents. 


## Exercise solutions

```{r fig.height=3, fig.width=4}
# Exerise 1 
## Make a histogram, Q-Q plot, and a box-plot for the life expectancy
## for a continent of your choice, but for all years. 
## Do the data appear normally distributed?

asia_plot = mydata %>%                          
  filter(continent %in% c("Asia"))              

p1 = asia_plot %>%                              
  ggplot(aes(x = lifeExp)) + 
  geom_histogram(bins = 15) #+                 
#facet_grid(. ~ year)                         # no facet

p2 = asia_plot %>%                          
  ggplot(aes(sample = lifeExp)) +               # `sample` for Q-Q plot
  geom_qq() + 
  geom_qq_line() #+ 
#facet_grid(. ~ year)                         # no facet

p3 = asia_plot %>%                              
  ggplot(aes(x = factor(year), y = lifeExp)) +  # year as factor
  geom_boxplot(aes(fill = factor(year))) +      
  geom_jitter(alpha = 0.4) +                    
  theme(legend.position = "none")               

library(patchwork)                              
p1 / p2 | p3
```

```{r fig.height=3, fig.width=4}
# Exercise 2
## Select any 2 years in any continent and perform a *t*-test to 
## determine whether mean life expectancy is significantly different. 
## Remember to plot your data first.

asia_years = mydata %>%                          
  filter(continent %in% c("Asia")) %>%          
  filter(year %in% c(1952, 1972))              

p1 = asia_years %>%                             
  ggplot(aes(x = lifeExp)) + 
  geom_histogram(bins = 15) +                 
  facet_grid(. ~ year)                          

p2 = asia_years %>%                            
  ggplot(aes(sample = lifeExp)) +              
  geom_qq() + 
  geom_qq_line() + 
  facet_grid(. ~ year)                        

p3 = asia_years %>%                             
  ggplot(aes(x = factor(year), y = lifeExp)) + 
  geom_boxplot(aes(fill = factor(year))) +      
  geom_jitter(alpha = 0.4) +                    
  theme(legend.position = "none")               

library(patchwork)                              
p1 / p2 | p3

asia_years %>% 
  t.test(lifeExp ~ year, data = .)
```

<!-- Changed to group_modify -->

```{r}
# Exercise 3
## In 2007, in which continents did mean life expectancy differ from 70
mydata %>% 
  filter(year == 2007) %>% 
  group_by(continent) %>%
  group_modify(
    ~ t.test(.$lifeExp, mu = 70) %>% tidy() # Somtimes awkward in the tidyverse
  )
```

<!-- Changed to group_modify -->
```{r}
# Exercise 4
## Use Kruskal-Wallis to determine if the mean population changed 
## significantly through the 1990s/2000s in individual continents. 

mydata %>% 
  filter(year >=1990) %>% 
  ggplot(aes(x = factor(year), y = pop)) + 
  geom_boxplot() + 
  facet_grid(. ~ continent)

mydata %>% 
  filter(year >=1990) %>% 
  group_by(continent) %>% 
  group_modify(
    ~ kruskal.test(pop ~ factor(year), data = .) %>% tidy()
  )
```
