# Working with categorical outcome variables

> Suddenly Christopher Robin began to tell Pooh about some of the things: People called Kings and Queens and something called Factors ... and Pooh said "Oh!" and thought how wonderful it would be to have a Real Brain which could tell you things.  
> A.A. Milne, *The House at Pooh Corner* (1928)

## Factors

We said earlier that continuous data can be measured and categorical data can be counted, which is useful to remember. 
Categorical data can be a:

* Factor
  + a fixed set of names/strings or numbers
  + these may have an inherent order (1st, 2nd 3rd) - ordinal factor
  + or may not (female, male) 
* Character
  + sequences of letters, numbers, and symbols
* Logical
  + containing only TRUE or FALSE

Health data is awash with factors. 
Whether it is outcomes like death, recurrence, or readmission. 
Or predictors like cancer stage, deprivation quintile, smoker yes/no.
It is essential therefore to be comfortable manipulating factors and dealing with outcomes which are categorical.

## The Question

We will use the classic “Survival from Malignant Melanoma” dataset which is included in the `boot` package. 
The data consist of measurements made on patients with malignant melanoma, a type of skin cancer. 
Each patient had their tumour removed by surgery at the Department of Plastic Surgery, University Hospital of Odense, Denmark between 1962 and 1977.

We are interested in the association between tumour ulceration and death from melanoma. 

## Get the data

The help page (`?boot::melanoma`) gives us the data dictionary.
This includes the defintion of each variable and the coding used.

```{r, message = F}
mydata = boot::melanoma
```

## Check the data
As always, check any new dataset carefully before you start analysis.  

```{r}
library(tidyverse)
library(finalfit)
mydata %>% glimpse()
mydata %>% ff_glimpse()
```

As can be seen, all of the variables are currently coded as continous/numeric.
The `<dbl>` stands for 'double', meaning numeric which comes from 'double-precision floating point', an awkward computer science term. 

## Recode the data
It is really important that variables are correctly coded for all plotting and analysis functions. 
Using the data dictionary, we will convert the categorical varibles to factors. 


```{r}
mydata = mydata %>% 
  mutate(sex.factor = 
           sex %>% 
           factor() %>% 
           fct_recode("Female" = "0",
                      "Male"   = "1") %>% 
           ff_label("Sex"),
         
         ulcer.factor = 
           ulcer %>% 
           factor() %>% 
           fct_recode("Present" = "1",
                      "Absent"  = "0") %>% 
           ff_label("Ulcerated tumour"),
         
         status.factor = 
           status %>% 
           factor() %>% 
           fct_recode("Died"  = "1",
                      "Alive" = "2",
                      "Died - other causes" = "3") %>% 
           ff_label("Status"))
```


## Should I convert a continuous variable to a categorical variable?













#the cut() function makes a continuous variable into a categorical variable
mydata$age %>% 
  cut(breaks = c(4,20,40,60,95), include.lowest=TRUE) ->
  mydata$age.factor


### Recap on factors

Press `F1` on `boot::melanoma` to see its description. 
Use the information from help to change the numbers into proper factors (e.g. 0 - female, 1 - male).

<!-- Dollars here which is quite late on, we should change this. Agree.  -->


## Chi-squared test / Fisher's exact test

### Plotting

Always plot new data first!

```{r, fig.width=4, fig.height=3}

mydata %>% 
	ggplot(aes(x = ulcer.factor, fill=status.factor)) + 
	  geom_bar(position = "fill") +
	  theme_bw() +
	  scale_fill_brewer(palette = "Paired")

```

```{r, fig.width=6, fig.height=3}

mydata %>% 
  ggplot(aes(x = age.factor, fill = status.factor)) +
    geom_bar() +
    theme_bw() +
    scale_fill_brewer(palette = "Paired")

```

```{r, fig.width=6, fig.height=3}

mydata %>% 
  ggplot(aes(x = ulcer.factor, fill=status.factor)) + 
    geom_bar() +
    theme_bw() +
    scale_fill_brewer(palette = "Paired") +
    facet_grid(sex.factor~age.factor)

```

## Analysis

### Using base R

First lets group together those that 'died of another cause' with those 'alive', to give a disease-specific mortality variable (`fct_collapse()` will help us). 

```{r}

mydata$status.factor %>%  
	fct_collapse("Alive" = c("Alive", "Died - other causes")) ->
  mydata$status.factor

```

Let's test mortality against sex.

```{r}

table(mydata$status.factor, mydata$sex.factor)
chisq.test(mydata$status.factor, mydata$sex.factor)

```

Note that `chisq.test()` defaults to the Yates' continuity correction. 

<!-- Lots of dollars here -->
It is fine to use this, but if you have a particular need not to, turn if off with `chisq.test(mydata$status.factor, mydata$sex.factor, correct=FALSE)`.

<!-- Crosstable in? -->
### Using `CrossTable`
This gives lots of useful information. 
It is readable in R and has lots of options, including Fisher's exact test. 
It is not that easy to extract results. 
\newpage
```{r}

library(gmodels)
# F1 CrossTable to see options
CrossTable(mydata$status.factor, mydata$sex.factor, chisq=TRUE)

```

### Exercise
Use the 3 methods (`table`, `chisq.test`, `CrossTable`) to test `status.factor` against `ulcer.factor`. 

```{r, eval=F}

table(mydata$status.factor, mydata$ulcer.factor)
chisq.test(mydata$status.factor, mydata$ulcer.factor)

```

Using `CrossTable`
```{r, eval=F}

CrossTable(mydata$status.factor, mydata$ulcer.factor, chisq=TRUE)

```

### Fisher's exact test

An assumption of the chi-squared test is that the 'expected cell count' is greater than 5. 
If it is less than 5 the test becomes unreliable and the Fisher's exact test is recommended. 

Run the following code. 

```{r}

library(gmodels)
CrossTable(mydata$status.factor, mydata$age.factor, expected=TRUE, chisq=TRUE)

```

Why does it give a warning? Run it a second time including `fisher=TRUE`. 


```{r}

library(gmodels)
CrossTable(mydata$status.factor, mydata$age.factor, expected=TRUE, chisq=TRUE)

```


## Summarising multiple factors (optional)
`CrossTable` is useful for summarising single variables. 
We often want to summarise more than one factor or continuous variable against our `dependent` variable of interest. 
Think of Table 1 in a journal article. 

## Summarising factors with `library(finalfit)`

This is our own package which we have written and maintain. 
It contains functions to summarise data for publication tables and figures, and to easily run regression analyses. 
We specify a `dependent` or outcome variable, and a set of `explanatory` or predictor varaibles. 

```{r, message = FALSE}

library(finalfit)
mydata %>% 
  summary_factorlist(dependent = "status.factor", 
                     explanatory = c("sex.factor", "ulcer.factor", "age.factor"),
                     p = TRUE,
                     column = TRUE)

```


### Summarising factors with `library(tidyverse)`

### Example

`Tidyverse` gives the flexibility and power to examine millions of rows of your data any way you wish. The following are intended as an extension to what you have already done. These demonstrate some more advanced approaches to combining `tidy` functions. 


```{r}

# Calculate number of patients in each group
counted_data = mydata %>%
  count(ulcer.factor, status.factor)

# Add the total number of people in each status group
counted_data2  = counted_data %>%
  group_by(status.factor) %>%
  mutate(total = sum(n))

```

```{r}

# Calculate the percentage of n to total
counted_data3 = counted_data2 %>%
  mutate(percentage = round(100*n/total, 1))

```

Create a combined columns of both `n` and `percentage` using `paste()` to add brackets around the percentage.

```{r}

counted_data4  = counted_data3 %>% 
  mutate(count_perc = paste0(n, " (", percentage, ")"))

```

Or combine everything together without the intermediate `counted_data` breaks.

```{r}

mydata %>%
  count(ulcer.factor, status.factor) %>%
  group_by(status.factor) %>%
  mutate(total = sum(n)) %>%
  mutate(percentage = round(100*n/total, 1)) %>% 
  mutate(count_perc = paste0(n, " (", percentage, ")")) %>% 
  select(-total, -n, -percentage) %>% 
  spread(status.factor, count_perc)

```

### Exercise

By changing one and only one word at a time in the above block (the "Combine everything together" section)

Reproduce this:

```{r, echo=F}

# Age
mydata %>%
  count(age.factor, status.factor) %>%
  group_by(status.factor) %>%
  mutate(total = sum(n)) %>%
  mutate(percentage = round(100*n/total, 1)) %>% 
  mutate(count_perc = paste0(n, " (", percentage, ")")) %>% 
  select(-total, -n, -percentage) %>% 
  spread(status.factor, count_perc) %>% 
  as.data.frame()

```

And then this:
```{r, echo=F}

# Sex
mydata %>%
  count(sex.factor, status.factor) %>%
  group_by(status.factor) %>%
  mutate(total = sum(n)) %>%
  mutate(percentage = round(100*n/total, 1)) %>% 
  mutate(count_perc = paste0(n, " (", percentage, ")")) %>% 
  select(-total, -n, -percentage) %>% 
  spread(status.factor, count_perc) %>% 
  as.data.frame()

```

Solution: The only thing you need to change is the first variable in `count()`, e.g., `count(age.factor, ...`.
