# (PART) Workflow {-}

Throughout this book we have tried to provide the most efficient approaches data analysis using R. 
In this section, we will provide workflows, or ways-of-working, which maximise efficiency, incorpirate reporting of results within analyses, make exporting of tables and plots easy, and keep data safe, secured and backed up. 
*consider a little more here*

# Notebooks and Markdown

## What is a Notebook?

R is all-powerful for the manipulation, visualisation and analysis of data. 
What is often be underappreciated is the flexibility in which analyses can be exported or reported.

For instance, a full scientific paper, industry report, or monthly update can be easily written to accommodate a varying underlying dataset, and all tables and plots will be updated automatically. 

This idea can be extended to a workflow in which all analyses are performed primarily within a document which represents doubles as the final report. 

Enter "Data Notebooks"!
Notebooks are documents which combine code and rich text elements, such as headings, paragraphs and links, in one document. 
They combine analysis and reporting in one human-readable document to provide an intuitive interface between the researcher and their analysis (Figure \@ref(fig:chap11-fig-literate). 
This sometimes gets called "literate programming", given the resulting logical structure of information can be easily read by human like a book.  

```{r chap11-fig-literate, echo = FALSE, fig.cap="CAPTION"}
knitr::include_graphics("images/chapter11/1_literate_programming.pdf")
```

In our own work, we have now moved to doing most of our analyses a Notebook environment, rather than a "script" file.  

Some of the advantages of the Notebook format are:

* code and output are adjacent to each other, so you are not constantly switching between "panes";
* easier to work on smaller screen;
* fully formated text elements can be included in documents;
* documentation and reporting can be done beside the code;
* the code itself can be outputted or hidden;
* faciliate collaboration by easily sharing human-readable analysis documents;
* can be outputted in a number of formats including html (web page), PDF, and Microsoft Word;
* output can be extended to other formats such as presentations;
* training/learning may be easier as course materials, examples, and student notes are all in same document.

## What is markdown?

Markdown is a lightweight language that can be used to write fully-formated documents. 
It is plaintext and uses a simple set of rules to produce rather sophisticated output - we love it!

It is easy to format headings, bold text, italics etc. 
Within RStudio there is a Quick Reference guide (Figure \@ref(fig:chap11-fig-help)) and links to the [RStudio cheatsheets](https://www.rstudio.com/resources/cheatsheets).

```{r chap11-fig-help, echo = FALSE, fig.cap="CAPTION"}
knitr::include_graphics("images/chapter11/3_help.pdf")
```

## What is the difference between a Notebook and a Markdown file?

The difference. 

## The anatomy of a notebook / R Markdown file

In RStudio a notebook can be easily created by going to File -> New File -> R Notebook.
A helpful template is provided to get you started. 
Figure \@ref(fig:chap11-fig-anatomy) shows the essential elements of a notebook file and how these translate to what is seen in the html preview. 

```{r chap11-fig-anatomy, echo = FALSE, fig.cap="CAPTION"}
knitr::include_graphics("images/chapter11/2_anatomy_rotated.pdf")
```

