# Exporting and reporting{#chap12-h1}

In chapter \ref(chap11-h1) we emphasised one of the less well known strengths of R  - the ease with which reports can be written in HTML (web page), PDF, or Word documents. 


```{r echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
mykable = function(x, caption = "CAPTION", ...){
  kable(x, row.names = FALSE, align = c("l", "l", "r", "r", "r", "r", "r", "r", "r"), 
        booktabs = TRUE, caption = caption, 
        linesep = c("", "", "\\addlinespace"), ...) %>%
    kable_styling(latex_options = c("scale_down", "hold_position"))
}
```

## The Question

What follows is for demonstration purposes and is not meant to illustrate model building. 
We will ask, does a particular characteristic of a colon cancer (differentiation) predict 5-year survival?


## Get the data

```{r message=FALSE}
library(tidyverse)
library(finalfit)
data(colon_s)
```

First explore variable of interest (exposure) by making it the dependent.

```{r, warning=FALSE, message=FALSE}
# Dependent variable
dependent = "differ.factor"

# Specify explanatory variables of interest
explanatory = c("age", "sex.factor", 
                "extent.factor", "obstruct.factor", 
                "nodes")
```

Note this useful alternative way of specifying explanatory variable lists:

```{r}
explanatory = colon_s %>% 
  select(age, sex.factor, extent.factor, obstruct.factor, nodes) %>% 
  names()
```

## Check the data

```{r, eval=FALSE}
colon_s %>% 
  ff_glimpse(dependent, explanatory)
```

## Demographics table

Look at associations between our exposure and other explanatory variables. 

```{r, eval=FALSE}
colon_s %>% 
  summary_factorlist(dependent, explanatory)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
colon_s %>% 
  summary_factorlist(dependent, explanatory, 
                     p=TRUE, na_include=TRUE) %>% 
  mykable(caption = "Exporting 'table 1': Tumour differentiation by patient and disease factors.")
```

Note missing data in `obstruct.factor` (REF missing data). 

We will drop this variable for now. 
Also see that nodes has not been labelled. 
There are small numbers in some variables generating `chisq.test()` warnings (expected fewer than 5 in any cell). 
Generate final table.

```{r, eval=FALSE}
colon_s = colon_s %>% 
  mutate(
    nodes = ff_label(nodes, "Lymph nodes involved"))

explanatory = c("age", "sex.factor", 
                "extent.factor", "nodes")

table1 = colon_s %>% 
  summary_factorlist(dependent, explanatory, 
                     p=TRUE, na_include=TRUE, 
                     add_dependent_label=TRUE
                     )
table1
```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
colon_s = colon_s %>% 
  mutate(
    nodes = ff_label(nodes, "Lymph nodes involved"))

explanatory = c("age", "sex.factor", 
                "extent.factor", "nodes")

colon_s %>% 
  summary_factorlist(dependent, explanatory, 
                     p=TRUE, na_include=TRUE, 
                     add_dependent_label=TRUE) %>% 
  mykable(caption = "Exporting 'table 1': adjusting labels and output.") %>% 
  column_spec(1, width = "3.5cm")
```

## Logistic regression table

Now examine explanatory variables against outcome. Check plot runs ok.

```{r, eval=FALSE}
explanatory = c("age", "sex.factor", 
                "extent.factor", "nodes", "differ.factor")
dependent = "mort_5yr"
table2 = colon_s %>% 
  finalfit(dependent, explanatory, 
           dependent_label_prefix = "")
table2
```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
explanatory = c("age", "sex.factor", 
                "extent.factor", "nodes", "differ.factor")
dependent = "mort_5yr"
colon_s %>% 
  finalfit(dependent, explanatory, 
           dependent_label_prefix = "") %>% 
  mykable(caption = "Exporting a regression results table.")
```

## Odds ratio plot

```{r fig.height=3.5, fig.width=7, message=FALSE, warning=FALSE, fig.cap="Odds ratio plot."}
colon_s %>% 
  or_plot(dependent, explanatory, 
          breaks = c(0.5, 1, 5, 10, 20, 30),
          table_text_size = 3.5)
```

## MS Word via knitr/R Markdown

Important. 
In most R Markdown set-ups, environment objects require to be saved and loaded to R Markdown document.

```{r, eval=FALSE}
# Save objects for knitr/markdown
save(table1, table2, dependent, explanatory, file = "out.rda")
```

We use RStudio Server Pro set-up on Ubuntu. But these instructions should work fine for most/all RStudio/Markdown default set-ups.

In RStudio, select File > New File > R Markdown.

A useful template file is produced by default. Try hitting knit to Word on the knitr button at the top of the .Rmd script window.

Now paste this into the file:

```` markdown
---
title: "Example knitr/R Markdown document"
author: "Your name"
date: "22/5/2020"
output:
word_document: default
---

`r ''````{r setup, include=FALSE}
# Load data into global environment. 
library(finalfit)
library(dplyr)
library(knitr)
load("out.rda")
```

## Table 1 - Demographics
`r ''````{r table1, echo = FALSE}
kable(table1, row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
```

## Table 2 - Association between tumour factors and 5 year mortality
`r ''````{r table2, echo = FALSE}
kable(table2, row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
```

## Figure 1 - Association between tumour factors and 5 year mortality
`r ''````{r figure1, echo = FALSE}
colon_s %>% 
or_plot(dependent, explanatory)
```
````

[The result is ok, but not great.](https://www.datasurg.net/wp-content/uploads/2018/05/example.docx) 


## Create Word template file

Now, edit your Word file to create a new template. Click on a table. The style should be `compact`. Right click > Modify... > font size = 9. Alter heading and text styles in the same way as desired. Save this as `template.docx`. Upload the file to your project folder. Add this reference to the .Rmd YAML heading, as below. Make sure you get the spacing correct.

The plot also doesn’t look quite right and it prints with warning messages. Experiment with fig.width to get it looking right.

Now paste this into your .Rmd file and run:

```` markdown
---
title: "Example knitr/R Markdown document"
author: "Ewen Harrison"
date: "21/5/2018"
output:
word_document:
reference_docx: template.docx  
---

`r ''````{r setup, include=FALSE}
# Load data into global environment. 
library(finalfit)
library(dplyr)
library(knitr)
load("out.rda")
```

## Table 1 - Demographics
`r ''````{r table1, echo = FALSE, results='asis'}
kable(table1, row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
```

## Table 2 - Association between tumour factors and 5 year mortality
`r ''````{r table2, echo = FALSE, results='asis'}
kable(table2, row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
```

## Figure 1 - Association between tumour factors and 5 year mortality
`r ''````{r figure1, echo = FALSE, warning=FALSE, message=FALSE, fig.width=10}
colon_s %>% 
or_plot(dependent, explanatory)
```
````

[This is now looking good, and further tweaks can be made if you wish.](https://www.datasurg.net/wp-content/uploads/2018/05/example2.docx)

## PDF via knitr/R Markdown

Default settings for PDF:

```` markdown
---
title: "Example knitr/R Markdown document"
author: "Ewen Harrison"
date: "21/5/2018"
output:
pdf_document: default
---

`r ''````{r setup, include=FALSE}
# Load data into global environment. 
library(finalfit)
library(dplyr)
library(knitr)
load("out.rda")
```

## Table 1 - Demographics
`r ''````{r table1, echo = FALSE, results='asis'}
kable(table1, row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
```

## Table 2 - Association between tumour factors and 5 year mortality
`r ''````{r table2, echo = FALSE, results='asis'}
kable(table2, row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
```

## Figure 1 - Association between tumour factors and 5 year mortality
`r ''````{r figure1, echo = FALSE}
colon_s %>% 
or_plot(dependent, explanatory)
```
````

[Again, this is not bad, but has issues.](https://www.datasurg.net/wp-content/uploads/2018/05/example.pdf)

We can fix the plot in exactly the same way. But the table is off the side of the page. For this we use the 'kableExtra` package. Install this in the normal manner. You may also want to alter the margins of your page using geometry in the preamble.

```` markdown
---
title: "Example knitr/R Markdown document"
author: "Ewen Harrison"
date: "21/5/2018"
output:
pdf_document: default
geometry: margin=0.75in
---

`r ''````{r setup, include=FALSE}
# Load data into global environment. 
library(finalfit)
library(dplyr)
library(knitr)
library(kableExtra)
load("out.rda")
```

## Table 1 - Demographics
`r ''````{r table1, echo = FALSE, results='asis'}
kable(table1, row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"),
booktabs=TRUE)
```

## Table 2 - Association between tumour factors and 5 year mortality
`r ''````{r table2, echo = FALSE, results='asis'}
kable(table2, row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"),
booktabs=TRUE) %>% 
kable_styling(font_size=8)
```

## Figure 1 - Association between tumour factors and 5 year mortality
`r ''````{r figure1, echo = FALSE, warning=FALSE, message=FALSE, fig.width=10}
colon_s %>% 
or_plot(dependent, explanatory)
```
````


[This is now looking pretty good for me as well.](https://www.datasurg.net/wp-content/uploads/2018/05/example2.pdf)

There you have it. A pretty quick workflow to get final results into Word and a PDF.
